Longitudinal Principal Manifold Estimation

### 2024-04-05 paper/main.tex
update discussion section

While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 297 participants, of whom 85 were considered cognitively normal, 150 displayed symptoms of mild cognitive impairment, and 62 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}). Of the 85 cognitively normal participants, 28 were subsequently diagnosed with mild cognitive impairment or dementia during the course of the study. Follow-up duration for these participants ranged from approximately 6 months to 112 months, with imaging scheduled to be conducted at six- or 12-month intervals depending on the stage of the study. To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow up were considered, meaning that the following analysis was conducted with data from 250 participants. Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.

@@

A second limitation is that the LPME method requires relative stability in the overall shape of the structure being fit over time. Thus, while this algorithm can be useful for separating meaningful changes in a manifold over time from inter-observation error, it will be unable to fit in situations where the error between observations is too substantial. How restrictive this requirement is depends largely on the area of application. When used to estimate subcortical surfaces as suggested here, this tends not to be overly restrictive, as changes in the shape and size of a subcortical structure, while clinically meaningful, tend to be relatively small in magnitude compared to their defining features.

Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset. It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus. This shortcoming could potentially be addressed in several ways. First, as discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated. Second, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated. Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}. The volume estimates obtained from the outputs of the PME and LPME algorithms reflected the insufficient fit of these algorithms to the hippocampus data. However, the volume estimation process includes ad hoc approaches to address missing data, and the high level of variability in the estimated volumes of the thalamus despite visual indications of improved fit to this data suggests that this approach may itself be contributing to this variability. Further testing of this approach will be needed before the LPME algorithm will be viable for approximating the volume of brain regions.

Additional work will also be necessary to more thoroughly explore the data augmentation approach proposed in this article and its capabilities. The simple calculations used here allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods. However, while the use of polar and spherical coordinates addressed a problem posed by well-behaved circular structures, it would not necessarily apply easily to the hippocampus or more complex structures, such as a figure 8 shape, or even rounded structures with large variations in the radius, which would result in several points sharing an angle from the origin. The development of a procedure to handle these considerations would greatly expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations. While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest.

>>>>

While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 297 participants, of whom 85 were considered cognitively normal, 150 displayed symptoms of mild cognitive impairment, and 62 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}). Of the 85 cognitively normal participants, 28 were subsequently diagnosed with mild cognitive impairment or dementia during the course of the study. Follow-up duration for these participants ranged from approximately 6 months to 112 months, with imaging scheduled to be conducted at six- or 12-month intervals depending on the stage of the study. To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow up were considered, meaning that the following analysis was conducted with data from 250 participants. Of these remaining participants, 80 were cognitively normal, 126 had mild cognitive impairment, and 44 were diagnosed with AD at their baseline visits. Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.

@@

A second limitation is that the LPME method requires relative stability in the overall shape of the structure being fit over time. Thus, while this algorithm can be useful for separating meaningful changes in a manifold over time from inter-observation error, it will be unable to fit in situations where the error between observations is too substantial. How restrictive this requirement is depends largely on the area of application. When used to estimate subcortical surfaces in a single individual as suggested here, this tends not to be overly restrictive, as changes in the shape and size of a subcortical structure, while clinically meaningful, tend to be relatively small in magnitude compared to their defining features.

Another limitation stems from the performance of the LPME method on the hippocampus surfaces obtained from the ADNI dataset. It is clear that neither the PME nor the LPME estimates were able to fully capture the distinctive curved structure of the hippocampus. There are two potential sources of this difficulty, each suggesting different steps that may be taken to address this issue. First, it is possible that this problem is caused by an inability of the PME and LPME algorithms to handle structures with as many sharp curves as the hippocampus. This may be addressed in several ways. As discussed above, the manifold learning method used in the initialization step of the PME algorithm, or the parameters of that method could be adjusted to better meet the needs of the specific manifold being estimated. Also, while a simple grid search is used to select the optimal value of the smoothing parameter in both the PME and LPME algorithms, it is possible that an improved search method would yield results that are more sensitive to the nuances of the structure being estimated. Finally, it may be possible to manipulate the manifold to be easier to estimate through adjustments to the data augmentation process, as described in Section \ref{s:LPME}.

The second potential source of the algorithms' poor fit to the hippocampus data could be an insufficiency in the data augmentation approach that allows us to fit some self-intersecting manifolds. The simple calculations used here, augmenting with polar or spherical coordiantes, allowed the PME and LPME algorithms to estimate self-intersecting manifolds that would not previously have been estimable by these methods. However, while this approach addressed a problem posed by well-behaved circular or spherical structures, such as the thalamus, its success may not extend to settings with more complex data. In the case of the hippocampus, we speculate that there may be intersections present in the spherical coordinates, which then result in the overlaps in certain regions of the estimated manifolds. Similar types of issues would arise in settings with other complex structures, like a figure 8 shape. The development of a procedure to handle these considerations would expand the potential use cases for this approach, thus allowing principal manifold-based methods to be used in a wider range of situations.

Further attention should be paid to the process used to estimate volumes from the PME and LPME outputs if these volume estimates are to be used as outcomes of interest. The current procedure includes ad hoc adjustments to maintain the stability of estimates in the presence of gaps in the estimated surface. However, alternate approaches may yield less variable results. For example, \cite{mengPrincipalManifoldEstimation2021} introduced an interior identification approach that effectively makes use of the explicit representation of the embedding map obtained through the PME and LPME fitting process, and which may be adapted to settings with gaps in the estimated manifold.

Finally, while the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, the further extension of this method to manifold estimation for groups of similar structures would also be of interest. Reaching estimates associated with groups of structures raises the possibility of statistical comparisons between these groups, which, among others, could prove beneficial in a number of clinical trial settings.
